---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(mxsem)
```

# mxsem

**mxsem** provides a **lavaan**-like (Rosseel, 2012) syntax to implement structural equation models (SEM)
with **OpenMx** (Boker et al., 2011). The objective is to simplify fitting basic SEM with **OpenMx** and
to allow users to easily switch between **lavaan** and **OpenMx**. The latter can
be useful if one of the two packages reports non-convergence or other issues.
Additionally, **mxsem** allows for parameter transformations and definition
variables.

> **Warning**: The syntax and settings of **mxsem** may differ from 
**lavaan** in some cases. See `?mxsem::mxsem` for more details on the syntax and the 
default arguments.

## Installation

The newest version of the package can be installed from GitHub using the following commands in R:

```{r, eval = FALSE}
if(!require(devtools)) install.packages("devtools")
devtools::install_github("jhorzek/mxsem", 
                         ref = "main")
```

## Example

The following example is directly adapted from **lavaan**:

```{r, eval = TRUE, results = 'hide', message=FALSE, warning=FALSE, message=FALSE, warning=FALSE}
library(mxsem)
model <- '
  # latent variable definitions
     ind60 =~ x1 + x2 + x3
     dem60 =~ y1 + a1*y2 + b*y3 + c1*y4
     dem65 =~ y5 + a2*y6 + b*y7 + c2*y8

  # regressions
    dem60 ~ ind60
    dem65 ~ ind60 + dem60

  # residual correlations
    y1 ~~ y5
    y2 ~~ y4 + y6
    y3 ~~ y7
    y4 ~~ y8
    y6 ~~ y8
'

mxsem(model = model,
      data  = OpenMx::Bollen) |>
  mxTryHard() |>
  summary()
```
<details>
<summary>Show summary</summary>


```{r, include=FALSE}
fit_mx <- mxsem(model = model,
      data  = OpenMx::Bollen) |>
  mxTryHard()
```
```{r,echo=FALSE}
summary(fit_mx)
```

</details>

## Adding bounds

Lower and upper bounds can be added to any of the parameters in the model. 
The following demonstrates bounds on a loading:
```{r, eval = TRUE, results = 'hide', message=FALSE, warning=FALSE}
library(mxsem)
model <- '
  # latent variable definitions
     ind60 =~ x1 + x2 + x3
     dem60 =~ y1 + a1*y2 + b*y3 + c1*y4
     dem65 =~ y5 + a2*y6 + b*y7 + c2*y8
     
     # lower bound on a1
     a1 > 0
     # upper bound on a2
     a2 < 10.123
'

mxsem(model = model,
      data  = OpenMx::Bollen, 
      # use latent variances to scale the model
      scale_loadings = FALSE, 
      scale_latent_variances = TRUE) |>
  mxTryHard() |>
  summary()
```

<details>
<summary>Show summary</summary>

```{r, include=FALSE}
fit_mx <- mxsem(model = model,
      data  = OpenMx::Bollen, 
      # use latent variances to scale the model
      scale_loadings = FALSE, 
      scale_latent_variances = TRUE) |>
  mxTryHard()
```
```{r,echo=FALSE}
summary(fit_mx)
```

</details>

**mxsem** adds lower bounds to any of the variances by default. To remove these
lower bounds, set `lbound_variances = FALSE` when calling `mxsem()`.

## Definition Variables

Definition variables are, for instance, used in latent growth curve models when
the time intervals between observations are different for the subjects
in the data set. Here is an example, where the variables `t_1`-`t_5` indicate
the person-specific times of observation:

```{r, eval = TRUE}
library(mxsem)
set.seed(3489)
dataset <- simulate_latent_growth_curve(N = 100)
head(dataset)
```

In **OpenMx**, parameters can be set to the values found in the columns of
the data set with the `data.` prefix. This is used in the following to fix
the loadings of a latent slope variable on the observations to the times 
recorded in `t_1`-`t_5`:
```{r, eval = TRUE, results = 'hide', message=FALSE, warning=FALSE}
library(mxsem)
model <- "
  I =~ 1*y1 + 1*y2 + 1*y3 + 1*y4 + 1*y5
  S =~ data.t_1 * y1 + data.t_2 * y2 + data.t_3 * y3 + data.t_4 * y4 + data.t_5 * y5

  I ~ int*1
  S ~ slp*1
  "

mxsem(model = model,
      data = dataset,
      add_intercepts = FALSE) |>
  mxTryHard() |>
  summary()
```

<details>
<summary>Show summary</summary>

```{r, include=FALSE}
fit_mx <- mxsem(model = model,
      data = dataset,
      add_intercepts = FALSE) |>
  mxTryHard()
```
```{r,echo=FALSE}
summary(fit_mx)
```

</details>

## Transformations

Sometimes, one may want to express one parameter as a function of other parameters.
For instance, in moderated non-linear factor analysis, model parameters are often 
expressed in terms of a covariate k. For instance, the effect $a$ of $\xi$ on $\eta$ 
could be expressed as $a = a_0 + a_1\times k$.

```{r, eval = TRUE}
library(mxsem)
set.seed(9820)
dataset <- simulate_moderated_nonlinear_factor_analysis(N = 100)
head(dataset)
```

```{r, eval = TRUE, results = 'hide', message=FALSE, warning=FALSE}
model <- "
xi =~ x1 + x2 + x3
eta =~ y1 + y2 + y3
eta ~ a*xi

# we need two new parameters: a0 and a1. These are created as follows:
!a0
!a1
# Now, we redefine a to be a0 + k*a1, where k is found in the data
a := a0 + data.k*a1
"

mxsem(model = model,
      data = dataset) |>
  mxTryHard() |>
  summary()
```

<details>
<summary>Show summary</summary>

```{r, include=FALSE}
fit_mx <- mxsem(model = model,
      data = dataset) |>
  mxTryHard()
```
```{r,echo=FALSE}
summary(fit_mx)
```

</details>

## Adapting the Model

`mxsem` returns an `mxModel` object that can be adapted further by users familiar
with **OpenMx**.


## References

* Boker, S. M., Neale, M., Maes, H., Wilde, M., Spiegel, M., Brick, T., Spies, J., Estabrook, R., Kenny, S., Bates, T., Mehta, P., & Fox, J. (2011).
OpenMx: An Open Source Extended Structural Equation Modeling Framework. Psychometrika, 76(2), 306–317. https://doi.org/10.1007/s11336-010-9200-6
* Rosseel, Y. (2012). lavaan: An R package for structural equation modeling. Journal of Statistical Software, 48(2), 1–36. https://doi.org/10.18637/jss.v048.i02
