---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(mxsem)
```

# mxsem

**mxsem** provides a **lavaan**-like (Rosseel, 2012) syntax to implement structural equation models (SEM)
with **OpenMx** (Boker et al., 2011). The objective is to simplify fitting basic SEM with
**OpenMx**, while also unlocking some very useful advanced features. For instance, 
**mxsem** allows for parameter transformations and definition
variables. However, **mxsem** is intentionally incomplete in order to focus
on simplicity. The main function (`mxsem()`) is similar to **lavaan**'s `sem()`-function
in that it tries to set up parts of the model automatically (e.g., add variances
automatically or scale the latent variables automatically).

> **Warning**: The syntax and settings of **mxsem** may differ from 
**lavaan** in some cases. See `vignette("Syntax", package = "mxsem")` for more details 
on the syntax and the default arguments.

## Alternatives

**mxsem** is not the first package providing a **lavaan**-like syntax for **OpenMx**.
You will find similar functions in the following packages:

- [**metaSEM**](https://github.com/mikewlcheung/metasem) provides a `lavaan2RAM`
function that can be combined with the `create.mxModel` function. This combination
offers more features than **mxsem**. For instance, constraints of the form `a < b`
are supported. In **mxsem** such constraints require algebras (e.g., `!diff; a := b - exp(diff)`).
- [**umx**](https://github.com/tbates/umx) provides the `umxRAM` and `umxLav2RAM` 
functions that can parse single **lavaan**-style statements (e.g., `eta =~ y1 + y2 + y3`) 
or an entire **lavaan** models to **OpenMx** models.
- [**tidySEM**](https://github.com/cjvanlissa/tidySEM) provides a unified syntax to
specify both, **lavaan** and **OpenMx** models. Additionally, it works well with the
**tidyverse**.
- [**ezMx**](https://github.com/OpenMx/ezMx) simplifies fitting SEM with **OpenMx**
and also provides a translation of **lavaan** models to **OpenMx** with the 
`lavaan.to.OpenMx` function.

## Installation

The newest version of the package can be installed from GitHub using the following commands in R:

```{r, eval = FALSE}
if(!require(devtools)) install.packages("devtools")
devtools::install_github("jhorzek/mxsem", 
                         ref = "main")
```

## Example

The following example is directly adapted from **lavaan**:

```{r, eval = TRUE, results = 'hide', message=FALSE, warning=FALSE, message=FALSE, warning=FALSE}
library(mxsem)
model <- '
  # latent variable definitions
     ind60 =~ x1 + x2 + x3
     dem60 =~ y1 + a1*y2 + b*y3 + c1*y4
     dem65 =~ y5 + a2*y6 + b*y7 + c2*y8

  # regressions
    dem60 ~ ind60
    dem65 ~ ind60 + dem60

  # residual correlations
    y1 ~~ y5
    y2 ~~ y4 + y6
    y3 ~~ y7
    y4 ~~ y8
    y6 ~~ y8
'

mxsem(model = model,
      data  = OpenMx::Bollen) |>
  mxTryHard() |>
  summary()
```
<details>
<summary>Show summary</summary>


```{r, include=FALSE}
fit_mx <- mxsem(model = model,
      data  = OpenMx::Bollen) |>
  mxTryHard()
```
```{r,echo=FALSE}
summary(fit_mx)
```

</details>

## Adding bounds

Lower and upper bounds can be added to any of the parameters in the model. 
The following demonstrates bounds on a loading:
```{r, eval = TRUE, results = 'hide', message=FALSE, warning=FALSE}
library(mxsem)
model <- '
  # latent variable definitions
     ind60 =~ x1 + x2 + x3
     dem60 =~ y1 + a1*y2 + b*y3 + c1*y4
     dem65 =~ y5 + a2*y6 + b*y7 + c2*y8
     
  # lower bound on a1
     a1 > 0
  # upper bound on a2
     a2 < 10.123
'

mxsem(model = model,
      data  = OpenMx::Bollen, 
      # use latent variances to scale the model
      scale_loadings = FALSE, 
      scale_latent_variances = TRUE) |>
  mxTryHard() |>
  summary()
```

<details>
<summary>Show summary</summary>

```{r, include=FALSE}
fit_mx <- mxsem(model = model,
      data  = OpenMx::Bollen, 
      # use latent variances to scale the model
      scale_loadings = FALSE, 
      scale_latent_variances = TRUE) |>
  mxTryHard()
```
```{r,echo=FALSE}
summary(fit_mx)
```

</details>

**mxsem** adds lower bounds to any of the variances by default. To remove these
lower bounds, set `lbound_variances = FALSE` when calling `mxsem()`.

## Definition Variables

Definition variables are, for instance, used in latent growth curve models when
the time intervals between observations are different for the subjects
in the data set. Here is an example, where the variables `t_1`-`t_5` indicate
the person-specific times of observation:

```{r, eval = TRUE}
library(mxsem)
set.seed(3489)
dataset <- simulate_latent_growth_curve(N = 100)
head(dataset)
```

In **OpenMx**, parameters can be set to the values found in the columns of
the data set with the `data.` prefix. This is used in the following to fix
the loadings of a latent slope variable on the observations to the times 
recorded in `t_1`-`t_5`:
```{r, eval = TRUE, results = 'hide', message=FALSE, warning=FALSE}
library(mxsem)
model <- "
  # specify latent intercept
     I =~ 1*y1 + 1*y2 + 1*y3 + 1*y4 + 1*y5
  # specify latent slope
     S =~ data.t_1 * y1 + data.t_2 * y2 + data.t_3 * y3 + data.t_4 * y4 + data.t_5 * y5
    
  # specify means of latent intercept and slope
     I ~ int*1
     S ~ slp*1
  
  # set intercepts of manifest variables to zero
     y1 ~ 0*1; y2 ~ 0*1; y3 ~ 0*1; y4 ~ 0*1; y5 ~ 0*1;
  "

mxsem(model = model,
      data = dataset) |>
  mxTryHard() |>
  summary()
```

<details>
<summary>Show summary</summary>

```{r, include=FALSE}
fit_mx <- mxsem(model = model,
      data = dataset) |>
  mxTryHard()
```
```{r,echo=FALSE}
summary(fit_mx)
```

</details>

## Transformations

Sometimes, one may want to express one parameter as a function of other parameters.
For instance, in moderated non-linear factor analysis, model parameters are often 
expressed in terms of a covariate k. For instance, the effect $a$ of $\xi$ on $\eta$ 
could be expressed as $a = a_0 + a_1\times k$.

```{r, eval = TRUE}
library(mxsem)
set.seed(9820)
dataset <- simulate_moderated_nonlinear_factor_analysis(N = 100)
head(dataset)
```

```{r, eval = TRUE, results = 'hide', message=FALSE, warning=FALSE}
model <- "
  # loadings
     xi =~ x1 + x2 + x3
     eta =~ y1 + y2 + y3
  # regression
     eta ~ a*xi
  
  # we need two new parameters: a0 and a1. These are created as follows:
     !a0
     !a1
  # Now, we redefine a to be a0 + k*a1, where k is found in the data
     a := a0 + data.k*a1
"

mxsem(model = model,
      data = dataset) |>
  mxTryHard() |>
  summary()
```

<details>
<summary>Show summary</summary>

```{r, include=FALSE}
fit_mx <- mxsem(model = model,
      data = dataset) |>
  mxTryHard()
```
```{r,echo=FALSE}
summary(fit_mx)
```

</details>

## Adapting the Model

`mxsem` returns an `mxModel` object that can be adapted further by users familiar
with **OpenMx**.

## Trouble shooting

Sometimes things may go wrong. One way to figure out where **mxsem** messed up
is to look at the parameter table generated internally. This parameter table
is not returned by default. See `vignette("create_parameter_table", package = "mxsem")`
for more details.


## References

* Boker, S. M., Neale, M., Maes, H., Wilde, M., Spiegel, M., Brick, T., Spies, J., Estabrook, R., Kenny, S., Bates, T., Mehta, P., & Fox, J. (2011).
OpenMx: An Open Source Extended Structural Equation Modeling Framework. Psychometrika, 76(2), 306–317. https://doi.org/10.1007/s11336-010-9200-6
* Rosseel, Y. (2012). lavaan: An R package for structural equation modeling. Journal of Statistical Software, 48(2), 1–36. https://doi.org/10.18637/jss.v048.i02
